---
title: Data visualization with ggplot2
author: Data Carpentry contributors
minutes: 60
---



```{r setup, echo=FALSE, purl=FALSE}
source("setup.R")
```


### Data Visualization with ggplot2

------------

> ### Learning Objectives
>
> * Produce scatter plots and histograms using ggplot.
> * Produce boxplots to summarize data.
> * Modify the aesthetics of an existing ggplot plot (including axis labels and color).
> * Build complex and customized plots from data in a data frame.
> * Set universal plot settings.

--------------

We start by loading the required packages. **`ggplot2`** is included in the **`tidyverse`** package.

```{r load-tidyverse-package, message=FALSE, purl=FALSE}
library(tidyverse)
```

Later we will learn how to read in data in R. For now we will compare the mutation rates between samples, starting by loading a table that combines SNP information from multiple samples.

```{r load-SNPs-tidy}
SNPs_tidy <- read_tsv("data/SNPs_tidy.txt")
SNPs_tidy
```                         


## Plotting with **`ggplot2`**

**`ggplot2`** is a powerful plotting package that makes it simple to create
complex plots from data in a data frame. It provides a programmatic interface
for specifying what variables to plot, how they are displayed, and general
visual properties. Therefore, we only need small changes to the code if the
underlying data change or if we decide to change from a bar plot to a
scatterplot. This interface helps in creating publication quality plots with
minimal amounts of adjustments and tweaking.

**`ggplot2`** functions like data in the 'long' tidy format, i.e., a column for
every dimension, and a row for every observation. Well-structured data will save
you lots of time when making figures with **`ggplot2`**. We start learning R by using **`ggplot2`** because it relies on concepts that we will need when we talk about data transformation in the next lessons.

ggplot graphics are built step by step by adding new elements. Adding layers in
this fashion allows for extensive flexibility and customization of plots.

To build a ggplot, we will use the following basic template that can be used for different types of plots:

```r
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) + <GEOM_FUNCTION>()
```

- use the `ggplot()` function and bind the plot to a specific data frame using the `data` argument

```{r, eval=FALSE, purl=FALSE}
ggplot(data = SNPs_tidy)
```

- define a mapping (using the aesthetic (`aes`) function), by selecting the variables to be plotted and specifying how to present them in the graph, e.g. as x/y positions or characteristics such as size, shape, color, etc.

```{r, eval=FALSE, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x = POS, y = REF))
```

- add 'geoms' – graphical representations of the data in the plot (points,
  lines, bars). **`ggplot2`** offers many different geoms; we will use some 
  common ones today, including:
  
      * `geom_point()` for scatter plots, dot plots, etc.
      * `geom_boxplot()` for, well, boxplots!
      * `geom_line()` for trend lines, time series, etc.  

To add a geom to the plot use the `+` operator. Because we have two continuous variables, let's use `geom_point()` first to create a scatter plot:

```{r first-ggplot, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x = POS, y = REF)) +
    geom_point()
```

The `+` in the **`ggplot2`** package is particularly useful because it allows
you to modify existing `ggplot` objects. This means you can easily set up plot
templates and explore different types of plots, so the above plot can also be
generated with code like this:

```{r, first-ggplot-with-plus, eval=FALSE, purl=FALSE}
## Assign plot to a variable
SNPs_plot <- ggplot(data = SNPs_tidy, aes(x = POS, y = REF))
## Draw the plot
SNPs_plot + 
    geom_point()
```

## Building your plots iteratively

<!-- the instructor guide should explain that for this section, the instructor should modify the same line of code instead of retyping each iteration -->

Building plots with **`ggplot2`** is typically an iterative process. We start by
defining the dataset we'll use, lay out the axes, and choose a geom:

```{r create-ggplot-object, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x = POS, y = REF)) +
    geom_point()
```

Now, we can start modifying this plot to extract more information from it. We can  add colors for all the points:

```{r adding-colors, purl=FALSE}					
ggplot(data = SNPs_tidy, aes(x = POS, y = REF)) +
    geom_point(color = "blue")
```

Or to color each alternate nucleotide substitution in the plot differently, you could specify the variable to use to the `color` argument within `aes()`:


```{r color-by-alternative, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x = POS, y = REF)) +
    geom_point(aes(color = ALT))
```

In this case **`ggplot2`** will automatically assign one color for each species.

Now at least we learnt some biology! G>C mutations are rare compared to G>T or G>A. 

> ### Challenge
>
> Use `shape` to differentiate reference nucleotide.

> ### Challenge
>
> Create a visualization that shows if variant call quality (QUAL) varies with 
> reference nucleotide
> 
> - Use QUAL as the y variable
> 
> Does quality also change with alternative variant?
> 
> - Colour the points by alternative nucleotide

## Density plots

What if we want to see the density of these across the genome (POSITION)?

GGplot provides histogram and smoothed density plots to enable this.

```{r pos-density, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x=POS)) +
    geom_histogram()
```

It is often helpful to control the width of histogram bins:

```{r pos-density-narrow, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x=POS)) +
    geom_histogram(binwidth = 100000)
```

> ### Challenge
> 
> Make a smoothed kernel density plot using `geom_density`.
> 
> - Can you change the smoothing width using the bandwidth `bw` argument?
> - Can you colour by mutation type (MUT)?
>


## Bar plots

What if we just want to count the different substitutions and display in a bar plot? The `geom_bar` command will help with this.

To count the number of substitutions with a given reference:

```{r barplot-ref, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x = MUT)) +
    geom_bar()
```

Now, to count the number of substitutions with a given reference and alternate combination, we could try to use color:

```{r barplot-color-by-alternative, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x = MUT, color = SAMPLE)) +
    geom_bar()
```

In fact, the `color` aesthetic governs the outside of the bar, and `fill` is needed to fill the inside:

```{r barplot-fill-by-alternative, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x = MUT, fill = SAMPLE)) +
    geom_bar()
```

To plot the bars dodged next to each other instead of stacked on top, we use the `position=position_dodge()`.

```{r barplot-fill-by-alternative-dodge, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x = MUT, fill = SAMPLE)) +
    geom_bar(position=position_dodge())
```


> ### Challenge
>
> Can you make SAMPLE the x-axis, and MUT the fill variable? 
> Which graph is more informative?
>


## Preventing points being plotted on top of each other

The position functions are very useful for avoiding plotting things on top of each other. A scatterplot of quality scores for each mutation type is here:

```{r quality, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x = MUT, y=QUAL)) +
    geom_point()
```

Many points are on top, we can't see the distribution. We can fix this with scatterplots with 'position="jitter"`. 

```{r quality-jitter, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x = MUT, y=QUAL)) +
    geom_point(position=position_jitter())
```

This `jitters` the points by a random amount in the x-direction, to make them visually separate. We can make better plots by controlling exactly how we jitter, for example by the `width` argument.

```{r quality-jitter-thin, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x = MUT, y=QUAL)) +
    geom_point(position=position_jitter(width=0.1))
```

> ### Challenge: removing overplotting
>
> Which jitter width gives the clearest plot?
>

## Boxplots to summarize data

Another way to summarize data is the boxplot

```{r quality-boxplot, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x = MUT, y=QUAL)) +
    geom_boxplot()
```

We can change colour and fill for a better-looking boxplot:

```{r quality-boxplot-blue, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x = MUT, y=QUAL)) +
    geom_boxplot(color="blue",fill=NA)
```

You can also plot multiple aesthetics on top of each other, so both a scatter plot and a boxplot.

```{r quality-boxplot-point, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x = MUT, y=QUAL)) +
    geom_point(position=position_jitter()) + 
    geom_boxplot(color="blue",fill=NA)
```

> ### Challenge: changing boxplot appearance
>
> Can you stop the outlier points be plotted by boxplot?
>
> - Try ?geom_boxplot for the help file, and search for "hiding the outliers" for instructions
>
> Can you make it easier to see the boxplot over the points?
> 
> - Try changing the `colour` or `alpha` of the points.
> - Try changing the line `size` of the boxplot.
> 
> Can you change the width of the boxplot *and* the position_jitter to match?
>

## Faceting

**`ggplot2`** has a special technique called *faceting* that allows the user to
split one plot into multiple plots based on discrete variables included in the
dataset.

Let's try the position histogram again

```{r pos-density-again, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x=POS)) +
    geom_histogram()
```

Now we can facet by sample, to get the count in each sample:

```{r pos-density-facet, purl=FALSE}
ggplot(data = SNPs_tidy, aes(x=POS)) +
    geom_histogram() +
    facet_wrap(~SAMPLE)
```

This again makes it clear that the number and position of mutations varies hugely between samples. 



> ### Challenge
>
> Can you facet by mutation type? 
> 
> Can you stack all plots on top of each other to compare position?
>
> - Try the `ncol` argument to facet_wrap
> - Or try `facet_grid`
> 


## Customizing your plot

### **`ggplot2`** themes

In addition to `theme_bw()`, which changes the plot background to white, **`ggplot2`**
comes with several other themes which can be useful to quickly change the look
of your visualization. The complete list of themes is available
at <http://docs.ggplot2.org/current/ggtheme.html>. `theme_minimal()` and
`theme_light()` are popular, and `theme_void()` can be useful as a starting
point to create a new hand-crafted theme.

The
[ggthemes](https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html)
package provides a wide variety of options (including an Excel 2003 theme). The
[**`ggplot2`** extensions website](https://www.ggplot2-exts.org) provides a list
of packages that extend the capabilities of **`ggplot2`**, including additional
themes.

There are many other options to change the plot appearance

### Changing the axes

* xlab
* ylab
* ggtitle

### Changing the legend

* labels
* legend position

> ### Challenge: plot appearance
>
> Can you change the appearance of your plot?
>
> - change overall appearance by adding theme_bw()?
> - change the x-axis label with xlab?
> - give your plot a title with ggtitle?
> 


## Arranging and Exporting 

* Show how to use ggsave, save files to `fig_output/` folder


<!--- things that could be added:

- geom_smooth() to show how to fit linear smoothing/linear lines
- geom_bar()/geom_col() and stacking/dodging
- difference between color and fill

--->



### More resources

* [ggthemeassist](https://github.com/calligross/ggthemeassist) is a RStudio-Addin that uses the [`rstudioapi`](https://github.com/rstudio/rstudioapi) package and provides a GUI for editing **`ggplot2`** themes.
* [ggedit](https://github.com/metrumresearchgroup/ggedit) is an R package that is used to facilitate **`ggplot2`** formatting. Aesthetic details can be refined easily.

```{r, child="_page_built_on.Rmd"}
```


```{r, eval=FALSE,purl=FALSE}
read_SRRvcf <- function(Sample="SRR097977",
                        prefix="data/",
                        suffix="_final_variants.vcf",
                        col_names=c("CHROM","POS","ID","REF","ALT",
                                    "QUAL","FILTER","INFO","FORMAT","VALUES")) {
    paste0(prefix,Sample,suffix) %>%
        read_tsv(col_names=col_names,comment="#") %>%
        mutate(SAMPLE=Sample)
        
}

variants_tidy <- list("SRR097977","SRR098026","SRR098027","SRR098028","SRR098281","SRR098283") %>%
    lapply(read_SRRvcf) %>%
    bind_rows

write_tsv(variants_tidy,"data/variants_tidy.txt")

add_mutationtype <- function(vdf,
                             mutkey=c("A>T"="AT>TA",
                                      "A>C"="AT>CG",
                                      "A>G"="AT>GC",
                                      "C>T"="CG>TA",
                                      "C>A"="CG>AT",
                                      "C>G"="CG>GC",
                                      "T>A"="AT>TA",
                                      "T>G"="AT>CG",
                                      "T>C"="AT>GC",
                                      "G>A"="CG>TA",
                                      "G>T"="CG>AT",
                                      "G>C"="CG>GC"),
                             mutlevels=c("AT>TA",
                                      "AT>CG",
                                      "AT>GC",
                                      "CG>TA",
                                      "CG>AT",
                                      "CG>GC")) {
    vdf %>% 
        mutate(MUT=paste(REF,ALT,sep=">") %>%
                   mutkey[.] %>%
                   factor(levels=mutlevels))
}

SNPs_tidy <- variants_tidy %>%
    filter(REF %in% single_nucs, ALT %in% single_nucs) %>%
    add_mutationtype %>%
    select(-ID,-FILTER)

write_tsv(SNPs_tidy,"data/SNPs_tidy.txt")
```

